---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# genderBR

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/genderBR)](https://cran.r-project.org/package=genderBR)
[![Travis-CI Build Status](https://travis-ci.org/meirelesff/genderBR.svg?branch=master)](https://travis-ci.org/meirelesff/genderBR)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/meirelesff/genderBR?branch=master&svg=true)](https://ci.appveyor.com/project/meirelesff/genderBR)
[![Package-License](https://img.shields.io/badge/license-GPL%20%28%3E=%202%29-brightgreen.svg?style=flat)](http://www.gnu.org/licenses/gpl-2.0.html)

`genderBR` predicts gender from Brazilian first names using data from the Instituto Brasileiro de Geografia e Estatistica's 2010 Census [API](http://censo2010.ibge.gov.br/nomes/).

## How does it work?

`genderBR`'s main function is `get_gender`, which takes a string with a Brazilian first name and predicts its gender using data from the IBGE's 2010 Census. More specifically, it extracts the number of female and male persons with a given first name and tries to predict the gender...

```{r}
library(genderBR)

get_gender("joão")
get_gender("ana")
```

Multiple names can be passed at the same function call:

```{r}
get_gender(c("pedro", "maria"))
```

And both full names and names written in lower or upper case are accepted as inputs:

```{r}
get_gender("Mario da Silva")
get_gender("ANA MARIA")
```

Additionally, one can filter results by state with the argument `state`; or get the probability that a given first name belongs to a female person by setting the `prob` argument to `TRUE` (defaults to `FALSE`).

```{r}
# What is the probability that the name Ariel belongs to a female person in Brazil?
get_gender("Ariel", prob = TRUE)

# What about differences between Brazilian states?
get_gender("Ariel", prob = TRUE, state = "RJ") # RJ, Rio de Janeiro
get_gender("Ariel", prob = TRUE, state = "RS") # RS, Rio Grande do Sul
get_gender("Ariel", prob = TRUE, state = "SP") # SP, Sao Paulo
```

Note that a vector with states' abbreviations is a valid input for `get_gender` function, so this also works:

```{r}
name <- rep("Ariel", 3)
states <- c("rj", "rs", "sp")
get_gender(name, prob = T, state = states)
```

This can be useful also to predict the gender of different individuals living in different states:

```{r}
df <- data.frame(name = c("Alberto da Silva", "Maria dos Santos", "Thiago Rocha", "Paula Camargo"),
                 uf = c("AC", "SP", "PE", "RS"),
                 stringsAsFactors = FALSE
                 )

df$gender <- get_gender(df$name, df$uf)

df
```


### Brazilian state abbreviations

The `genderBR` package relies on Brazilian state abbreviations (acronyms) to filter results. To get a complete dataset with the full name, IBGE code, and abbreviations of all 27 Brazilian states, use the `get_states` functions:

```{r}
get_states()
```

## Geographic distribution of Brazilian first names

The `genderBR` package can also be used to get information on the relative and total number of persons with a given name by gender and by state in Brazil. To that end, use the `map_gender` function:

```{r}
map_gender("maria")
```

To specify gender in the consultation, use the optional argument `gender` (valid inputs are `f`, for female; `m`, for male; or `NULL`, the default option).

```{r}
map_gender("iris", gender = "m")
```

## Installing

`genderBR` is not yet available on CRAN. To install a development version, use:

```{r, eval = FALSE}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("meirelesff/genderBR")
```
    
## Data

The data used in this package come from the Instituto Brasileiro de Geografia e Estatistica's (IBGE) 2010 Census. The surveyed population includes 190,8 million Brazilians -- with more than 130,000 unique first names.

To extracts the numer of male and female persons with a given first name in Brazil, the package uses the [API](http://censo2010.ibge.gov.br/nomes/) provided by the IBGE. In this service, names with different spelling (e.g., Ana and Anna, or Marcos and Markos) are considered different names, and only names with more than 20 occurrences, or more than 15 occurrences in a given state, were included in the database.

For more information on the IBGE's data, please check (in Portuguese): 
http://censo2010.ibge.gov.br/nomes/

## Author

[Fernando Meireles](http://fmeireles.com)
    
