
---
title: 'genderBR: Predicting gender from Brazilian first names'
tags:
  - R
  - gender inference
  - Brazil
  - census data
authors:
  - name: Fernando Meireles
    orcid: 0000-0002-7027-2058
    affiliation: "1"
    corresponding: true
affiliations:
 - name: Instituto de Estudos Sociais e Politicos, Universidade do Estado do Rio de Janeiro, Brazil
   index: 1
date: 16 January 2026
bibliography: paper.bib
---

# Summary

Quantitative social science research frequently relies on large-scale administrative or scraped datasets that lack demographic indicators -- including gender. While alternatives to infer gender from first names are often used to circumvent the problem, existing tools mostly target Anglophone populations and use non-public, commercial data and methods. `genderBR` is an R package designed as an alternative to infer gender from Brazilian first names calculating name-by-gender probabilities using official data from the 2010 and 2022 Brazilian Censuses. The package offers a fast offline mode via an internal dataset and an online mode that queries IBGE's API and supports state-level filters, allowing predictions to be calibrated to reflect both temporal and regional variations in name-gender associations. By using large scale census microdata, `genderBR` thus enables academics, journalists, and others to obtain binary gender probabilities to study aggregate demographic groups in a reproducible manner.

# Statement of need

A common challenge in quantitative social science is the lack of demographic information in large-scale datasets. For instance, researchers frequently utilize administrative records and other data sources that do not contain gender information to study topics as diverse as elections and descriptive representation [@colner2025running; @lucas2021women; @warshaw2022local]; labor market inequalities [@banerjee2025gender; @wu2020gender]; academic publishing patterns and career trajectories [@dion2018gendered; @gaule2018advisor; @hagan2020women; @liu2023gender; @mulders2024matter]; the effects of gender composition of jury trials [@flanagan2018race]; among many others. As these datasets often include first names, scholars often resort to imputing gender based on first names, exploiting naming conventions that correlate with a binary gender classification.

Despite the limitations of this approach, obtaining gender labels from first names remains a popular method in the absence of direct measures. For example, solutions such as @genderAPI, @Namsor, and @genderize offer commercial APIs that allows users to query first names and obtain gender predictions based on large proprietary datasets. In the open-source domain, packages such as `gender` [@Mullen2016; see also @shi2025open and @tzioumis2018demographic] rely on historical datasets from the United States or Europe, using counts of names by gender from sources like the US Social Security Administration to predict gender from first names. Yet, while useful for Anglophone contexts, these tools are not well-suited for the study of other populations due to cultural and linguistic differences in naming conventions. In a comparative assessment, @vanhelene2024inferring show that these methods are highly accurate for Western populations, with both @genderAPI and @genderize achieving accuracies exceeding 98% in classifying a global dataset of mostly Western names. However, their performances dropped below 82% for names from South Korea, China, Singapore, and Taiwan. The open-source alternative, the `gender` package [@Mullen2016], achieved an even lower overall accuracy, of only 79.8% using the IPUMS method and 85.7% using US Social Security Administration data. In a different study, @santamaria2018comparison corroborates these findings for Asian populations, showing also that, even among Western names, commercial solutions can leave as much as 20% of names unclassified due to non-coverage.

`genderBR` addresses theses gaps by offering an open-source solution to tackle this problem specifically for Brazil, whose population exceeds 200 million people and represents the largest Portuguese-speaking country in the world. Soecifically, the package is grounded in Brazilian Institute of Geography and Statistics (IBGE) census microdata [@IBGE2022], which provides and standardizes Brazilian first names occurrence counts by gender -- coded a binary classification based on self-reported data collected in the census. The package supports both national and state-level predictions for 2010 and 2022, providing probabilities alongside hard labels. Differently from API-based solutions that rely on private or non-representative scraped data, `genderBR` leverages official census to offer almost universal coverage of Brazilian names^[According to the official documentation from IBGE [@IBGE2022], only first names with at least 5 occurrences in the census are included in the public microdata. However, as the package combines data from two different census, this threshold still allows for a comprehensive coverage of common names.], allowing accurate and reproducible gender assignment for Brazilian populations.

# Method and Usage

`get_gender` is the package's core function: it cleans names (lowercases, strips accents, keeps only the first token), aggregates duplicates to reduce computation time, and returns either binary labels or female-use probabilities. Users can switch between census years (`year = 2010` or `2022`), request probabilities (`prob = TRUE`), tune decision thresholds (`threshold`, default is 0.9), and choose the data source. When `internal = TRUE`, the default, and no state is provided, results come from a bundled probability table (`nomes`), enabling fully offline and fast analyses using the R package @datatable as backend for fast lookups. When a state is supplied or `internal = FALSE`, the function queries IBGE's API and, if the input vector is large, inserts small pauses to respect rate limits. Below are some usage examples.

## Examples

```{r, message=FALSE, include=FALSE}
options(digits = 4)
library(genderBR)
```

After installing and loading the package, users can call the main function as follows. To obtain national predictions with binary labels for a vector of names, user can pass a vector of Brazilian first or full names to the `get_gender` function:

```{r}
# Return labels
name_vec <- c("João", "Ana Maria da Silva", "alex")
get_gender(name_vec)

# Return probabilities of given names being female
get_gender(name_vec, prob = TRUE)
```

To obtain the probabilities calibrated to specific period and Brazilian states, users can provide a `state` vector with the corresponding two-letter state abbreviations (e.g., "RJ" for Rio de Janeiro), in which case will query IBGE's API directly instead of using the internal data to calculate name-by-gender-state probabilities. Important note: the length of the `state` vector must match that of the `name` vector. 

```{r}
# Probabilities for "Ariel"
states <- get_states()
name <- rep("Ariel", nrow(states))
states$prob_fem10 <- get_gender(name, prob = TRUE, state = states$abb, year = 2010)
states$prob_fem22 <- get_gender(name, prob = TRUE, state = states$abb, year = 2022)

head(states)
```

Internally, `genderBR` computes $p_\text{female} = \frac{f}{f + m}$ for each name (and state, when provided), where $f$ and $m$ are IBGE female and male counts. The returned label applies the rule: female if $p_\text{female} > \tau$, male if $p_\text{female} < 1 - \tau$, and unknown otherwise, with the default $\tau = 0.9$. This decision treshold is deterministic, but users can tune $\tau$ to trade off precision and coverage for their applications. For example:

```{r}
# Different thresholds
get_gender("Darcy", threshold = 0.5)
get_gender("Darcy", threshold = 0.7)
```

# Ethical Considerations

While `genderBR` provides a practical solution for imputing gender from first names, it is crucial to acknowledge the implications of using this approach. By relying on a binary gender classification derived from naming conventions recorded at birth, the provided method is unable to differentiate between non-binary gender identities or changes in gender identity over time. In line with recommendations from similar packages [@Mullen2016], users should avoid using `genderBR` to impose binary classifications to individuals or in contexts where misclassification may lead to harm or discrimination against groups. Instead, the package should be regarded as an estimator for aggregate, large populations -- to approximate the proportion of female partisan affiliates in the whole country, for example. In sum, `genderBR` should be viewed as a last resort tool when self-identified gender data is lacking and infering it from first names does not pose risks to groups under study.

# Acknowledgements

I thank the Brazilian Institute of Geography and Statistics for providing the public census name data and API that make this work possible, and the `R` and academic community that contributed to the development of the package with feedback and suggestions.

# References
